Suitability prevalence area index in late Quaternary explains genetic diversity in Tassel-eared Squirrels

### 2023-07-10 manuscript_english.tex
Updates from Overleaf

\title{Suitability prevalence area index in late quaternary explains
genetic diversity in Tassel eared Squirrels}
\author{Norma Hernandez, Angel Robles Nathan Upham}
\date{2023-06-27}

@@

As a case study we consider squirrels (\emph{Sciurus aberti}), a species
currently distributed in disjunct patches from the southern Rocky
Mountains in the United States to the northern Sierra Madre Occidental
in Mexico for which it is possible to find reliable information on both
the genetic structure of its populations and its current distribution
(Bono et al. 2018; Burgin et al. 2018).

Our results reveal that fitness prevalence corresponds to the fixation
index of \emph{S. aberti} populations with respect to a source
population. Populations closer to the historical endemic area present a
higher genetic diversity and a lower \(F_{st}\) value. Finally, this
study allows us to add a biogeographic explanation to the results
obtained with population genetic methods and to generate maps of this
structure as tools to support conservation with a perspective that
integrates both population genetics and historical patterns of species
distribution.

@@

The current distributions of species do not always correspond to their
historical distributions over evolutionarily significant periods. This
is because environmental conditions are not static over time; species
tend to distribute where conditions are most favorable.

@@

We define \(\vec{e}_i =(e_1, e_2, ..., e_i )\) as a vector of
environmental variables, that is, a point in the space of \(v\)
environmental variables. Then \(R_i(\vec{e}_i)\) is a function that
relates fitness to environment, that is, for a species with
non-overlapping generations, the population's net reproductive rate
\(R_i\) is a measure of fitness that depends on environmental
conditions.

That is, the sites in the geography \(\mathbf{G}_o\) that present
positive growth rates are related to favorable conditions. If we take
the set of places in ambient space where these values are greater than a
threshold, we can define a volume in multidimensional space such that
\(R(\vec{e})\) is always positive and greater than this threshold, that
is \(\mathbf{N}_F(t) = \{\vec{e} | R_0(\vec{e}(t)) > k \}\).

This is a way of expressing the fundamental niche as a function of
environmental conditions. So we can map each point of geography to
ambient space. Thus, for each site in the geography there will
correspond a site in the environmental space with different
environmental conditions and different growth rates. However all sites
within the region defined by \(\mathbf{N}_F\) will have viable
populations and positive rates. However, we assume that the environment
is dynamic and each geographic point is static, so for each site in
geographic space this growth rate depends on the environment and
therefore is different at each time.

Thus locally the growth rate will be related to the local population
size as long as it is within the boundary defined by \(\mathbf{N}_F\).
Therefore, the local population size as a function of the intrinsic rate
that depends on the change in environmental conditions over time is:
\(N_i(t+1) = R_0(\vec{e_i}(t) ) N_i(t)\). That is, the locally effective
population size is a function of ideal environmental conditions.

We start by considering this defined volume of the set of \(R_0\) as a
set of trajectories for each point from geographic space to
environmental space. Considering this rate fixed \(\mathbf{R}_0\) when
the population grows, that is,
\(\mathbf{N}_F(t) = \{\vec{e} | R_0(\vec{e}(t)) > 1 \}\) and therefore
any local population will have a positive growth rate.

@@

This equation was proposed by Soberon and Peterson Soberón and Peterson
(2020) to relate demographics to a function of fitness and environment.
In that study they tested the validity with monthly variables and
empirical values on a short time scale. relating life history parameters
and the environmental suitability of niche models.

This set can be seen as a convex volume in environmental space and we
relate this growth rate to an environmental suitability function bounded
between \(0\) and \(1\), for example, for a unimodal multivariate pdf,
it would be equivalent to the distance from the maximum of the function.
That is, the fitness-related environmental suitability can be any
function such that it is \(0\) if \(R_i(\vec{e}_i) > 1\) and \(1\) when
\(R\) is evaluated at the expected value of \(\vec {e}\) for all values
of \(R_i(\vec{e}_i) > 1\). In a practical way, \(\mathbf{N}_F\) is
approximated from points of presence in the geography obtained from
databases, numerically approximating a convex surface with a surface
\(\mathbf{N}_R\) of the environmental space.

@@

The objective of this theoretical approach is not to know the shape of
the function of \(R_0\) but rather to find this local variation relating
this value of \(R_0\) and the function \(\mathbf{N}_F\) with a function
\$ S\_i(~vec\{e\}\_i)\$ of environmental suitability that can be
estimated from known methods of applied mathematics and machine learning
(for example the case of Maxent, Phillips, Anderson, and Schapire
(2006)).

Es decir:

>>>>

\title{Suitability prevalence area index in late Quaternary explains
genetic diversity in Tassel-eared Squirrels}
\author{Norma Hernandez, Angel Robles Nathan Upham}
\date{2023-06-27}

@@

As a case study we consider Tassel-earred Squirrels (\emph{Sciurus aberti}), a species
currently distributed in disjunct patches of pine forest from the southern Rocky
Mountains in the United States to the northern Sierra Madre Occidental
in Mexico for which reliable population-genetic and occurrence information is available. Our results reveal that the genetic fixation
index (\(F_{st}\)) of \emph{S. aberti} populations with respect to a source
population correspond to the area of greatest prevalence through time, as modeled over the last 120,000 years. Populations closer to the historical endemic area display a
higher genetic diversity and a lower \(F_{st}\) value. This result suggests a general biogeographic explanation for the common observation of geographic clines of genetic diversity within species. By generating a time series of habitat suitability maps for a given species,  the suitability prevalence area can be estimated. Thus, integrating  population-genetic data with estimates of historical patterns of species
distribution holds promise for improving conservation and management efforts.

@@

The current distributions of species may not correspond to their
historical distributions over evolutionary significant time periods. This
is because environmental conditions are not static over time, and the populations of a species
tend to 'track' environments, distributing where conditions are most favorable in the absence of barriers (REF).

@@

We define \(\vec{e}_i =(e_1, e_2, ..., e_n )\) as a vector of
environmental variables, that is, a point in the space of \(n\)
environmental variables. Then \(R_i(\vec{e}_i)\) is a function that
relates fitness to environment, that is, for a species with
non-overlapping generations, the population's net reproductive rate
\(R_i\) is a measure of fitness that depends on environmental
conditions.

That is, the sites in the geography \(\mathbf{G}_o\) that present
positive growth rates are related to favorable conditions. If we take
the set of places in ambient space where these values are greater than a
threshold, we can define a volume in multidimensional space such that
\(R(\vec{e})\) is always positive and greater than this threshold, that
is \(\mathbf{N}_F = \{\vec{e} | R_0(\vec{e}(t)) > k \}\).

This is a way of expressing the fundamental niche as a function of
environmental conditions. So we can map each point of geography to
ambient space. Thus, for each site in the geography there will
correspond a site in the environmental space with different
environmental conditions and different growth rates. However all sites
within the region defined by \(\mathbf{N}_F\) will have viable
populations and positive rates. We assume that the environment
is dynamic and each geographic point is static, so for each site in
geographic space this growth rate depends on the environment and
therefore is different at each time.

Thus locally the growth rate will be related to the local population
size as long as it is within the boundary defined by \(\mathbf{N}_F\).
Therefore, the local population size as a function of the intrinsic rate
that depends on the change in environmental conditions over time is:
\(N_i(t+1) = R_0(\vec{e}_i(t) ) N_i(t)\). That is, the locally effective
population size is a function of ideal environmental conditions.

We start by considering this defined volume of the set of \(R_0\) as a
set of trajectories for each point from geographic space to
environmental space. Considering this rate fixed \(R_0\) when
the population grows, that is,
\(\mathbf{N}_F(t) = \{\vec{e} | R_0(\vec{e}(t)) > 1 \}\) and therefore
any local population will have a positive growth rate.

@@

This equation was proposed by Soberon and Peterson
(2020) to relate demographics to a function of fitness and environment.
In that study they tested the validity with monthly variables and
empirical values on a short time scale. relating life history parameters
and the environmental suitability of niche models.

This set can be seen as a convex volume in environmental space and we
relate this growth rate to an environmental suitability function bounded
between \(0\) and \(1\), for example, for a unimodal multivariate probability density function,
it would be equivalent to the distance from the maximum of the function.
That is, the fitness-related environmental suitability can be any
function such that it is \(0\) if \(R_i(\vec{e}_i) > 1\) and \(1\) when
\(R\) is evaluated at the expected value of \(\vec {e}\) for all values
of \(R_i(\vec{e}_i) > 1\). In a practical way, \(\mathbf{N}_F\) is
approximated from points of presence in the geography obtained from
databases, numerically approximating a convex surface with a surface
\(\mathbf{N}_R\) of the environmental space.

@@

The objective of this theoretical approach is not to know the shape of
the function of \(R_0\) but rather to find this local variation relating
this value of \(R_0\) and the function \(\mathbf{N}_F\) with a function
\(S_i(\vec\{e\}\_i)\) of environmental suitability that can be
estimated from known methods of applied mathematics and machine learning
(for example the case of Maxent, Phillips, Anderson, and Schapire
(2006)):
